Flowers text analysis

### 2022-11-16 code/annotation_analysis.Rmd
work on draft cogsci figs, much sbert

library(here)
library(tidyverse)
library(rstanarm)
library(rstan)
library(tidybayes)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
#for plotting
theme_set(theme_bw())
#Data import constants
date_start=lubridate::ymd('2021-07-19')
fig_path=here("figs")
read_data_path <- "data/processed_data/joined_data/"
annotation_path <- "data/annotations"
model_path <- "code/models"

@@

# Analysis

```{r}
d.games <- read_csv(here(paste(read_data_path, "games.csv", sep = "")),
                      col_types = cols()) %>% distinct() |> 
    mutate(conditionName = condition, 
         condition = case_when(conditionName == "coopMulti" ~ "Shared Utilities",
                               conditionName == "competCartel" ~ "Individual Utilities",
                               conditionName == "coopCartel" ~ "Old Cooperative",
                               TRUE ~ conditionName)) |> 
  select(gameId,conditionName)

annotated <-  read_csv(here(annotation_path, "flowers_annotations.csv")) |> 
  filter(!is.na(span)) |> 
  mutate(numchar=str_length(span),
         numword=str_count(span, "\\W+") + 1) |> 
  left_join(d.games) |> 
  mutate(flower=str_c(referent,color),
         player=str_c(gameId, name)) |> 
  select(gameId, player, flower, trialNum, pResp, normalized_span,color,numchar, numword, conditionName)
```

## How many words of referring language are there per round?

@@

sum_words <- annotated %>% group_by(trialNum, gameId,conditionName) %>% summarize(sum_char=sum(numchar),
                                           sum_word=sum(numword))

  # ggplot(sum_words,aes(x=trialNum,y=sum_char, color=conditionName))+
  #   #geom_line(aes(group=gameId))+
  #   facet_grid(.~conditionName)+
  #   geom_smooth(aes(group=gameId), method="lm", se=F)+
  #   geom_jitter(alpha=.5)

    # ggplot(sum_words,aes(x=trialNum,y=sum_char, color=conditionName))+
    # geom_line(aes(group=gameId))+
    # facet_grid(.~conditionName)
    # 
    # 
      ggplot(sum_words,aes(x=trialNum,y=sum_word, color=conditionName))+
    #geom_line(aes(group=gameId))+
    facet_grid(.~conditionName)+
    geom_smooth(aes(group=gameId), method="lm", se=F)+
    geom_jitter(alpha=.5)

    # ggplot(sum_words,aes(x=trialNum,y=sum_word, color=conditionName))+
    # geom_line(aes(group=gameId))+
    # facet_grid(.~conditionName)
    #

```
There are reduction trends (not sure if linear fit is correct or not) for most groups, some serious group differences, also noise.

## How long are individual referring expressions?

 # ggplot(annotated,aes(x=trialNum,y=numchar))+ geom_jitter(alpha=.1)+ geom_smooth()+ facet_grid(.~conditionName)
  ggplot(annotated,aes(x=trialNum,y=numword))+geom_jitter(alpha=.1)+ geom_smooth()+ facet_grid(.~conditionName)

```
Individual referring expressions tend to get shorter.

## How many referring expressions are there?
```{r}

annotated |> group_by(gameId, trialNum, conditionName) |> tally() |> 
  ggplot(aes(x=trialNum, y=n))+geom_jitter()+geom_smooth()+facet_grid(.~conditionName)
# 
# annotated |> group_by(gameId, referent, conditionName) |> tally() |> 
#   group_by(conditionName) |> summarize(avg=mean(n), median=median(n))
```
Mostly the number of referring expressions is relatively constant, but the expressions themselves shorten.

@@

m1 <- stan_lmer(numword ~ trialNum + (1|gameId)+ (1|player)+(1|flower), data=annotated) |> write_rds(here(model_path,"m1.rds"))

@@

# Sbert stuff

@@

```{r, include=F,}
sbert_concat <- read_csv(here(annotation_path,"pre_sbert_concat.csv")) %>% bind_cols(readRDS(here(annotation_path,'post_sbert_concat.RData')))  %>% as_tibble()

```{r, include=F,}
d.games <- read_csv(here(paste(read_data_path, "games.csv", sep = "")),
                      col_types = cols()) %>% distinct() |> 
    mutate(conditionName = condition, 
         condition = case_when(conditionName == "coopMulti" ~ "Shared Utilities",
                               conditionName == "competCartel" ~ "Individual Utilities",
                               conditionName == "coopCartel" ~ "Old Cooperative",
                               TRUE ~ conditionName)) |> 
  select(gameId,conditionName)

sbert_single <- read_csv(here(annotation_path,"pre_sbert.csv")) %>% bind_cols(readRDS(here(annotation_path,'post_sbert.RData')))  %>% as_tibble() |> 
  left_join(d.games)

```{r, include=F}

F_mat <- sbert_single %>% select(starts_with("...")) %>% as.matrix() #Features
M_mat <- sbert_single %>% select(-starts_with("...")) %>% mutate(feature_ind=row_number()) |> 
  mutate(quarter=case_when(
    trialNum < 6 ~ 1,
    trialNum < 12 ~ 2,
    trialNum < 18 ~ 3,
    trialNum < 24 ~ 4
  ))
```

@@

## Do flower labels diversify within games?

Hypothesis: descriptions for different flowers get more different over time (within game)

@@

flower_distinctive <- M_mat %>% 
  group_by(gameId,quarter, conditionName) %>% 
  mutate(combinedId=referent) %>% 
  make_across_df(F_mat, 'cosine') %>% 
  rename(referent1=dim1,referent2=dim2) %>% 
  mutate(sim = ifelse(is.nan(sim), NA, sim)) %>%
  filter(referent1!=referent2) %>% 
  ungroup()
```

```{r}
ggplot(flower_distinctive, aes(x=quarter,y=sim,color=conditionName))+
      stat_summary(aes(group=str_c(gameId), color=conditionName),fun.data = "mean_cl_boot", position = position_dodge(width=.3), alpha=.5, geom="point")+geom_smooth(formula=y~poly(x,2))
```

## Do groups diverge on labels?

Hypothesis: descriptions for the same flower between games get more different over time (maybe, for some flowers)

@@

game_distinctive <- M_mat %>% 
  group_by(quarter, color, referent) %>% 
  mutate(combinedId=gameId) %>% 
  make_across_df(F_mat, 'cosine') %>% 
  rename(gameId1=dim1,gameId2=dim2) %>% 
  mutate(sim = ifelse(is.nan(sim), NA, sim)) %>%
  filter(gameId1!=gameId2) %>% 
  ungroup()
```

```{r}
ggplot(game_distinctive, aes(x=quarter,y=sim,color=color))+
      stat_summary(aes(group=str_c(color,referent), color=color),fun.data = "mean_cl_boot", position = position_dodge(width=.3), alpha=.5, geom="point")+geom_smooth(formula=y~poly(x,2))
```

Wide variability -- some flowers might get more consistent names than others?

## Do group members converge with each other?

Hypothesis: players within the same game will label the same flower more similarly later in the game than earlier.

@@

player_similar <- M_mat %>% 
  group_by(quarter, color, referent, gameId) %>% 
  mutate(combinedId=name) %>% 
  make_across_df(F_mat, 'cosine') %>% 
  rename(name1=dim1,name2=dim2) %>% 
  mutate(sim = ifelse(is.nan(sim), NA, sim)) %>%
  filter(name1!=name2) %>% 
  ungroup()
```

```{r}
ggplot(player_similar, aes(x=quarter,y=sim,color=color))+
      stat_summary(aes(group=str_c(color,referent), color=color),fun.data = "mean_cl_boot", position = position_dodge(width=.3), alpha=.5, geom="point")+geom_smooth(formula=y~poly(x,2))
```

>>>>

library(here)
library(tidyverse)
library(rstanarm)
library(rstan)
library(tidybayes)
library(viridis)
library(cowplot)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
#for plotting
theme_set(theme_bw())
#Data import constants
date_start=lubridate::ymd('2021-07-19')
fig_path=here("figs")
read_data_path <- "data/processed_data/joined_data/"
annotation_path <- "data/annotations"
model_path <- "code/models"

@@

# Prep

```{r}
d.games <- read_csv(here(paste(read_data_path, "games.csv", sep = "")),
                      col_types = cols()) %>% distinct() |> 
    mutate(conditionName = condition, 
         condition = case_when(conditionName == "coopMulti" ~ "Shared Utilities",
                               conditionName == "competCartel" ~ "Individual Utilities",
                               conditionName == "coopCartel" ~ "Old Cooperative",
                               TRUE ~ conditionName)) |> 
  select(gameId,conditionName, condition)

annotated <-  read_csv(here(annotation_path, "flowers_annotations.csv")) |> 
  filter(!is.na(span)) |> 
  mutate(numchar=str_length(span),
         numword=str_count(span, "\\W+") + 1) |> 
  left_join(d.games) |> 
  mutate(flower=str_c(referent,color),
         player=str_c(gameId, name)) |> 
  select(gameId, player, flower, trialNum, pResp, normalized_span,color,numchar, numword, condition)
```

# How much data

```{r, message=T}

message("total utterances")
annotated |> tally()

message("Utterances / game")

annotated |> group_by(gameId,condition) |> tally() |> pull(n) |> summary()

message("Utterances / flower")
annotated |> group_by(flower) |> tally() |> pull(n) |> summary()

# Word count basics

Take away -- across the condition manipulation, the number of utterances doesn't decrease, but the length of each one does.

Graph could use finessing if going in paper.

```{r, fig.height=4, fig.width=8}

sum_words <- annotated %>% group_by(trialNum, gameId,condition) %>% summarize(sum_char=sum(numchar),
                                           sum_word=sum(numword))

     total <-  ggplot(sum_words,aes(x=trialNum+1,y=sum_word, color=condition))+
   # geom_smooth(method="lm", se=T)+
            geom_smooth(method="lm", formula=y~poly(x,2), se=T)+

        geom_jitter(alpha=.1, width=.2, height=.5)+
        labs(y="Referential Words per trial per game", x="Trial Number", color="")+
        theme(legend.position="bottom", legend.title=element_blank())+
        scale_x_continuous(limits=c(1,24), breaks=c(6,12,18,24))+
        scale_color_brewer(palette="Dark2")

    legend <- get_legend(total)

    all_words <- total+theme(legend.position = "none")

        words <- ggplot(annotated,aes(x=trialNum,y=numword, color=condition))+geom_jitter(alpha=.1)+ geom_smooth()+
          labs(y="Words per reference expression", x="Trial Number")+
        theme(legend.position="none", legend.title=element_blank())+
        scale_x_continuous(limits=c(1,24), breaks=c(6,12,18,24))+
        scale_color_brewer(palette="Dark2")

        utts <- annotated |> group_by(gameId, trialNum, condition) |> tally() |> 
  ggplot(aes(x=trialNum, y=n, color=condition))+geom_jitter(alpha=.02)+geom_smooth()+
          labs(y="Reference expressions  per trial per game", x="Trial Number")+
        theme(legend.position="none", legend.title=element_blank())+
        scale_x_continuous(limits=c(1,24), breaks=c(6,12,18,24))+
        scale_color_brewer(palette="Dark2")

        top <- plot_grid(all_words, words, utts,nrow=1)
        plot_grid(top, legend, nrow=2, rel_heights=c(1,.1))

# How many different flowers described / trial?

We can probably give this as summary stats and not as graph

Per game

@@

annotated |> group_by(gameId, trialNum, condition) |> select(flower) |> unique() |> tally() |> 
  ggplot(aes(x=trialNum, y=n, color=condition))+geom_jitter(height=.25, alpha=.2)+geom_smooth()+
          labs(y="Distinct flowers referenced per trial", x="Trial Number")+
        theme(legend.position="none", legend.title=element_blank())+
        scale_x_continuous(limits=c(1,24), breaks=c(6,12,18,24))+
        scale_color_brewer(palette="Dark2")
```

Per person

```{r}
annotated |> group_by(gameId, trialNum, condition, player) |> select(flower) |> unique() |> tally() |> 
  ggplot(aes(x=trialNum, y=n, color=condition))+geom_jitter(height=.25, alpha=.1)+geom_smooth()+
          labs(y="Distinct flowers referenced per player per trial", x="Trial Number")+
        theme(legend.position="none", legend.title=element_blank())+
        scale_x_continuous(limits=c(1,24), breaks=c(6,12,18,24))+
        scale_color_brewer(palette="Dark2")
```

@@

m1 <- stan_lmer(numword ~ trialNum + condition + (1|gameId)+ (1|player)+(1|flower), data=annotated) |> write_rds(here(model_path,"m1.rds"))

@@

# Sbert notes

@@

```{r, include=F,}
d.games <- read_csv(here(paste(read_data_path, "games.csv", sep = "")),
                      col_types = cols()) %>% distinct() |> 
    mutate(conditionName = condition, 
         condition = case_when(conditionName == "coopMulti" ~ "Shared Utilities",
                               conditionName == "competCartel" ~ "Individual Utilities",
                               conditionName == "coopCartel" ~ "Old Cooperative",
                               TRUE ~ conditionName)) |> 
  select(gameId,conditionName, condition)

 d.raw_chat <- read_csv(here(paste(read_data_path, 
                                   "raw_chat.csv", 
                                   sep="")),
                       col_types = cols()) %>% distinct() |> 
   select(gameId, playerId) |> unique()

 d.players <- read_csv(here(paste(read_data_path, "players.csv", sep = "")),
                       col_types = cols()) %>% distinct() |> select(playerId,name) |> inner_join(d.raw_chat)

 sbert_single <- read_csv(here(annotation_path,"pre_sbert.csv"))|> 
    left_join(d.players) |> left_join(d.games) |> mutate(seen="game") |> 
   select(playerId, gameId, condition, text=normalized_span,referent, color, trialNum, seen) %>% bind_cols(readRDS(here(annotation_path,'post_sbert.RData')))  %>% as_tibble()

sbert_post_test <- read_csv(here(annotation_path,"post_test.csv")) |> left_join(d.games) |> 
  mutate(trialNum=25) |> 
  select(playerId, gameId, condition, text,referent=num, color, trialNum, seen)%>% bind_cols(readRDS(here(annotation_path,'post_sbert_post_test.RData')))  %>% as_tibble()

sbert_all <- sbert_single |> union(sbert_post_test)
F_mat <- sbert_all %>% select(starts_with("...")) %>% as.matrix() #Features
M_mat <- sbert_all %>% 
  select(-starts_with("...")) %>% 
  mutate(feature_ind=row_number(),
         sixth=case_when(
           trialNum < 4 ~ 1,
          trialNum < 8 ~ 2,
          trialNum < 12 ~ 3,
          trialNum < 16 ~ 4,
          trialNum < 20 ~ 5,
          trialNum < 24 ~ 6,
          trialNum > 24 ~ 7
  ))
```

@@

# Sbert pictures

These have quadratic smooths fit onto them, to allow for curvature, but no reason to choose this curve and not another.

## During game

@@

# by trial
flower_all <- M_mat %>% 
  filter(seen %in% c("game")) |> 
  group_by(color,condition) %>% 
  mutate(combinedId=str_c(referent,"_", trialNum, "_",gameId,"_",playerId)) %>% 
  make_across_df(F_mat, 'cosine') %>% 
  separate(dim1, c("ref1", "trialNum1","game1","player1"), convert=T) |> 
  separate(dim2, c("ref2", "trialNum2", "game2", "player2"), convert=T) |> 
 # rename(referent1=dim1,referent2=dim2) %>% 
  mutate(sim = ifelse(is.nan(sim), NA, sim)) %>%
  #filter(ref1!=ref2) %>% 
  mutate(earlier=ifelse(trialNum1>trialNum2,trialNum2, trialNum1) |> as.numeric(),
         later=ifelse(trialNum1>trialNum2, trialNum1,trialNum2) |> as.numeric()) |> 
  ungroup()
```

```{r}
flower_all|> 
  filter(later-earlier<3) |> 
  mutate(distance=case_when(
    player1==player2 ~ "Same person",
    game1==game2 ~ "same game",
    T ~ "different game"),
    same_flower=ifelse(ref1==ref2, "Same flower", "diff flower")
  ) |> 
ggplot( aes(x=later,y=sim,color=condition))+     
  stat_summary(aes(group=str_c(color,condition), color=condition),fun.data = "mean_cl_boot", position = position_dodge(width=.3), alpha=.3, geom="point")+
  geom_smooth(formula=y~poly(x,2))+facet_grid(same_flower~distance)+theme(legend.position="bottom", legend.title=element_blank())+
        scale_x_continuous(limits=c(1,24), breaks=c(6,12,18,24))+
        scale_color_brewer(palette="Dark2")

Comparing utterances within rounds that were up to 2 apart, round coded as the later (so 10-12,11-12 and 12-12 will all contribute to "12"). Up to 2 apart is somewhat arbitrary, not sure what the right rolling window to do is.

Take aways:
 - descriptions to the same flower within game (either same person or not) become more similar
 - descriptions to the same flower across games become (slightly) less similar
 - descriptions to different flowers become less similar (regardless of between/within game)

## Between game and end

We can treat what someone wrote at the end as the convention and then compare how similar this is to earlier utterances by them and their group mates.

Big question is whether the end utts need cleaning (which probably yes?)

We see that for the same flower, similarity to the end utt increases over blocks both within individual and indiv - other group mate (this is all within a game)

@@

# by trial
to_end <- M_mat %>% 
  filter(seen %in% c("game", "own")) |> 
  group_by(gameId,color,condition) %>% 
  mutate(combinedId=str_c(referent,"_", trialNum, "_",seen,"_",playerId)) %>% 
  make_across_df(F_mat, 'cosine') %>% 
  separate(dim1, c("ref1", "trialNum1","seen1","player1"), convert=T) |> 
  separate(dim2, c("ref2", "trialNum2", "seen2", "player2"), convert=T) |> 
 # rename(referent1=dim1,referent2=dim2) %>% 
  mutate(sim = ifelse(is.nan(sim), NA, sim)) %>%
  filter(seen1!=seen2) %>% 
  mutate(earlier=ifelse(trialNum1>trialNum2,trialNum2, trialNum1) |> as.numeric(),
         later=ifelse(trialNum1>trialNum2, trialNum1,trialNum2) |> as.numeric()) |> 
  ungroup()
```

Comparing utterances earlier to post-game descriptions.

General question of how to display these things -- what should be color v faceting and what to do for the dots.

```{r}
to_end|> 
  mutate(distance=case_when(
    player1==player2 ~ "Same person",
    T ~ "same game"),
        same_flower=ifelse(ref1==ref2, "Same flower", "diff flower")
) |> 
ggplot( aes(x=earlier,y=sim,color=condition))+     
  stat_summary(aes(group=str_c(color,condition), color=condition),fun.data = "mean_cl_boot", position = position_dodge(width=.3), alpha=.3, geom="point")+
  geom_smooth(formula=y~poly(x,2))+facet_grid(same_flower~distance)+theme(legend.position="bottom", legend.title=element_blank())+
        scale_x_continuous(limits=c(1,24), breaks=c(6,12,18,24))+
        scale_color_brewer(palette="Dark2")

##  Within end-stuff

We can look at how converged people are (or are not) by comparing what they say at the end.

Each person was asked for names ( ~ "how would you label this to your group") for the 12 images they were seeing and then for 4 from a different color palette.

Own = color you saw all game

Other = the color you didn't see

Mixed = own for one person, other for other person

@@

# by trial
in_end <- M_mat %>% 
  filter(seen %in% c("other", "own")) |> 
  group_by(color,condition) %>% 
  mutate(combinedId=str_c(referent, "_",seen,"_", gameId, "_",playerId)) %>% 
  make_across_df(F_mat, 'cosine') %>% 
  separate(dim1, c("ref1", "seen1","game1","player1"), convert=T) |> 
  separate(dim2, c("ref2",  "seen2","game2", "player2"), convert=T) |> 
 # rename(referent1=dim1,referent2=dim2) %>% 
  mutate(sim = ifelse(is.nan(sim), NA, sim)) %>%
  ungroup()
```

```{r}
in_end|> 
  mutate(distance=case_when(
    player1==player2 ~ "Same person",
    game1==game2 ~ "same game",
    T ~ "different game"),
    same_flower=ifelse(ref1==ref2, "Same flower", "diff flower"),
    same_color=case_when(
  seen1=="own" & seen2=="own" ~ "own",
  seen2=="other" & seen1=="other" ~ "other",
  T ~"mixed"
  ))|> 
ggplot( aes(x=same_color,y=sim,color=condition))+     
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width=.3))+
facet_grid(same_flower~distance)+theme(legend.position="bottom", legend.title=element_blank())+
        scale_color_brewer(palette="Dark2")
```

Not sure how to show spread here. Or how to do a viz that allows for comparisons.

Take aways:

- People use more similar descriptions for two flowers that weren't in their set than those that were (checks out -- in set ones have been described and diverged)
- This same thing holds (with overall lower sims) for group-mates
- Group mates are more similar in how they describe flowers they've seen (v not) (at least in shared utils)
- not sure how to interpret the diff game -- this is the baseline? Maybe it's just that you're more similar for the same flower versus different regardless? (also this is baseline for other panels)

# Future things

- have we tried log space? 
- Could do more end game -- earlier comparisons
- tSNE??
- Models of (some of ?) what's shown in forgoing graphs
- CLIP ????
- clean up end utts? 
- are there interesting lang -- performance connections ?

## Boring diligence
- Double check that non-talkers aren't in this data at all (end game joins). Consider coding and grouping end stuff by what color they usually saw during the game
- do sbert a different way and confirm (split v aggregate?)
